---
title: "<strong>Multilevel regression and post-stratification</strong>"
output: 
  html_document:
    code_folding: show
    theme: paper
    toc: yes
    toc_float: 
      collapsed: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = "", fig.align = "center", 
                      comment = "", cache = TRUE)
```

```{css, echo=FALSE}
blockquote {
    padding: 10px 20px;
    margin: 0 0 20px;
    font-size: 14px;
    color: #828282;
    border-left: 10px solid #EEE;
}

body {
    font-size: 14px;
}
```

<p align="right"><font size="2">
See [___here___](https://mc-stan.org/rstanarm/articles/mrp.html) for the official vignette by Lauren Kennedy and Jonah Gabry. This notebook is simply an adaptation of what is written there.
</font><p>

****

TO DO: FIND GELMAN'S TEXTBOOK AND READ UP ON THIS PART.


```{r, message=FALSE}
## Packages
library(tidyverse)
library(rstanarm)
options(mc.cores = parallel::detectCores())

## Plot settings
custom_theme <- function (base_family = "Roboto", ...) {
    theme_minimal(...) %+replace% 
    theme(
      plot.title = element_text(face = "bold", hjust = 0), 
      panel.grid.minor = element_blank(), 
      panel.grid.major = element_line(size = 0.2), 
      strip.background = element_rect(fill = "gray80", color = "gray80")
      )
}

theme_set(custom_theme())
```

```{r, include=FALSE, collapse=TRUE}
simulate_mrp_data <- function(n) {
  J <- c(2, 3, 7, 3, 50) # male or not, eth, age, income level, state
  poststrat <- as.data.frame(array(NA, c(prod(J), length(J)+1))) # Columns of post-strat matrix, plus one for size
  colnames(poststrat) <- c("male", "eth", "age","income", "state",'N')
  count <- 0
  for (i1 in 1:J[1]){
    for (i2 in 1:J[2]){
      for (i3 in 1:J[3]){
        for (i4 in 1:J[4]){
          for (i5 in 1:J[5]){
              count <- count + 1
              # Fill them in so we know what category we are referring to
              poststrat[count, 1:5] <- c(i1-1, i2, i3,i4,i5) 
          }
        }
      }
    }
  }
  # Proportion in each sample in the population
  p_male <- c(0.52, 0.48)
  p_eth <- c(0.5, 0.2, 0.3)
  p_age <- c(0.2,.1,0.2,0.2, 0.10, 0.1, 0.1)
  p_income<-c(.50,.35,.15)
  p_state_tmp<-runif(50,10,20)
  p_state<-p_state_tmp/sum(p_state_tmp)
  poststrat$N<-0
  for (j in 1:prod(J)){
    poststrat$N[j] <- round(250e6 * p_male[poststrat[j,1]+1] * p_eth[poststrat[j,2]] *
      p_age[poststrat[j,3]]*p_income[poststrat[j,4]]*p_state[poststrat[j,5]]) #Adjust the N to be the number observed in each category in each group
  }
  
  # Now let's adjust for the probability of response
  p_response_baseline <- 0.01
  p_response_male <- c(2, 0.8) / 2.8
  p_response_eth <- c(1, 1.2, 2.5) / 4.7
  p_response_age <- c(1, 0.4, 1, 1.5,  3, 5, 7) / 18.9
  p_response_inc <- c(1, 0.9, 0.8) / 2.7
  p_response_state <- rbeta(50, 1, 1)
  p_response_state <- p_response_state / sum(p_response_state)
  p_response <- rep(NA, prod(J))
  for (j in 1:prod(J)) {
    p_response[j] <-
      p_response_baseline * p_response_male[poststrat[j, 1] + 1] *
      p_response_eth[poststrat[j, 2]] * p_response_age[poststrat[j, 3]] *
      p_response_inc[poststrat[j, 4]] * p_response_state[poststrat[j, 5]]
  }
  people <- sample(prod(J), n, replace = TRUE, prob = poststrat$N * p_response)
  
  ## For respondent i, people[i] is that person's poststrat cell,
  ## some number between 1 and 32
  n_cell <- rep(NA, prod(J))
  for (j in 1:prod(J)) {
    n_cell[j] <- sum(people == j)
  }
  
  coef_male <- c(0,-0.3)
  coef_eth <- c(0, 0.6, 0.9)
  coef_age <- c(0,-0.2,-0.3, 0.4, 0.5, 0.7, 0.8, 0.9)
  coef_income <- c(0,-0.2, 0.6)
  coef_state <- c(0, round(rnorm(49, 0, 1), 1))
  coef_age_male <- t(cbind(c(0, .1, .23, .3, .43, .5, .6),
                           c(0, -.1, -.23, -.5, -.43, -.5, -.6)))
  true_popn <- data.frame(poststrat[, 1:5], cat_pref = rep(NA, prod(J)))
  for (j in 1:prod(J)) {
    true_popn$cat_pref[j] <- plogis(
      coef_male[poststrat[j, 1] + 1] +
        coef_eth[poststrat[j, 2]] + coef_age[poststrat[j, 3]] +
        coef_income[poststrat[j, 4]] + coef_state[poststrat[j, 5]] +
        coef_age_male[poststrat[j, 1] + 1, poststrat[j, 3]]
      )
  }
  
  #male or not, eth, age, income level, state, city
  y <- rbinom(n, 1, true_popn$cat_pref[people])
  male <- poststrat[people, 1]
  eth <- poststrat[people, 2]
  age <- poststrat[people, 3]
  income <- poststrat[people, 4]
  state <- poststrat[people, 5]
  
  sample <- data.frame(cat_pref = y, 
                       male, age, eth, income, state, 
                       id = 1:length(people))
  
  #Make all numeric:
  for (i in 1:ncol(poststrat)) {
    poststrat[, i] <- as.numeric(poststrat[, i])
  }
  for (i in 1:ncol(true_popn)) {
    true_popn[, i] <- as.numeric(true_popn[, i])
  }
  for (i in 1:ncol(sample)) {
    sample[, i] <- as.numeric(sample[, i])
  }
  list(
    sample = sample,
    poststrat = poststrat,
    true_popn = true_popn
  )
}
```

## Introduction

Multilevel regression and post-stratification (MRP) is a method for adjusting your sample so that it becomes more representative of the population (for a set of key variables).


Data

>As this is a simple toy example, we will describe the proportion of the population who would choose to adopt a cat over a dog, given the opportunity. We will simulate data using a function that is included in the appendix of this document. The simulate_mrp_data() function simulates a sample from a much larger population. It returns a list including the sample, population poststratification matrix and the true population preference for cats.

```{r}
mrp_sim <- simulate_mrp_data(n = 1200) 
str(mrp_sim)  
```

- `true_popn`

- `poststrat`

- `sample`

The following chunk saves these three data frames to the global environment.

```{r}
list2env(mrp_sim, envir = globalenv())
```


>The variables describing the individual (age, ethnicity, income level and gender) will be used to match the sample to the population of interest. To do this we will need to form a post-stratification table, which contains the number of people in each possible combination of the post-stratification variables. We have 4 variables with 2 (male), 7 (age), 3 (ethnicity) and 3 (income) levels, so there are 2x7x3x3 different levels. Participants are also selected from a state (50), increasing the number of possible levels to 6300.



TBH, I really don't get it....


```{r}
sample$type <- "sample"
true_popn$type <- "population"

df <- bind_rows(sample, true_popn)

ggsummary <- function(df, var) {
  
  x_format <- pull(df, {{var}}) %>% unique() %>% sort()

  df %>% 
    group_by({{var}}, type) %>% 
    summarize(cat_pref = mean(cat_pref)) %>% 
    ggplot(aes({{var}}, cat_pref, linetype = type, group = type)) + 
    geom_point() + geom_line() +
    scale_x_continuous(breaks = x_format, labels = x_format) + 
    scale_y_continuous(labels = scales::percent) + 
    theme(legend.position = "top") + 
    labs(linetype = NULL)

}

ggsummary(df, age)
ggsummary(df, eth)
ggsummary(df, income)
ggsummary(df, male)
```


